Fractional Types

### 2013-01-29 frac.tex
frac update
fixed macros
fixed typo
reorganization

%subcode-inline{bnf-inline} name langRev
%! swap+ = \mathit{swap}^+
%! swap* = \mathit{swap}^*
%! dagger =  ^{\dagger}
%! assocl+ = \mathit{assocl}^+
%! assocr+ = \mathit{assocr}^+
%! assocl* = \mathit{assocl}^*
%! assocr* = \mathit{assocr}^*
%! identr* = \mathit{uniti}
%! identl* = \mathit{unite}
%! dist = \mathit{distrib}
%! factor = \mathit{factor}
%! (o) = \fatsemi
%! (;) = \fatsemi
%! (*) = \times
%! (+) = +

%subcode-inline{bnf-inline} regex \{\{(((\}[^\}])|[^\}])*)\}\} name main include langRev
%! [^ = \ulcorner
%! ^] = \urcorner
%! [v = \llcorner
%! v] = \lrcorner
%! |-->* = \mapsto^{*}
%! |-->> = \mapsto_{\ggg}
%! |-->let = \mapsto_{let}
%! |--> = \mapsto
%! <--| = \mapsfrom
%! |- = \vdash
%! in = \!\!\in\!\!
%! <=> = \Longleftrightarrow
%! <-> = \leftrightarrow
%! ~> = \leadsto
%! ::= = ::=
%! /= = \neq
%! vi = v_i
%! di = d_i
%! si = s_i
%! sj = s_j
%! F = \texttt{F}
%! T = \texttt{T}
%! forall = \forall
%! exists = \exists
%! empty = \emptyset
%! eta = \eta
%! where = \textbf{where}
%! epsilon = \varepsilon
%! least = \phi
%! trace+ = \mathit{trace}^{+}
%! trace* = \mathit{trace}^{\times}
%! loop+ = loop_{+}
%! loop* = loop_{\times}
%! CatC = {\mathcal C}
%! CatA = {\mathcal A}
%! gamma = \gamma
%! {[ = \{
%! ]} = \}
%! elem = \in
%! dagger = ^\dagger
%! alpha = \alpha
%! beta = \beta
%! rho = \rho
%! @@ = \mu
%! @ = \,@\,
%! langRev = \Pi
%! langRevT = \Pi^{o}
%! langRevEE = \Pi^{\eta\epsilon}_{+}
%! bullet = \bullet
%! * = \times

@@

\begin{document}
\title{Fractional Types}
\author{}
\institute{}
\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
?
\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Syntax and Semantics}

We can turn the above system of isomorphisms into a (reversible)
programming language by simply providing a syntax for each of the
isomorphisms. For each isomorphism, we introduce two
\emph{combinators} corresponding to the left-to-right and
right-to-left directions.

@@

\noindent To summarize the syntax of our language {{langRev}} is as follows.

\begin{definition}[Syntax of {{langRev}}]
\label{def:langRev}
We collect our types, values, and combinators, to get the full language
definition:
%subcode{bnf} include main
% value types, b ::= 0 | 1 | b+b | b*b 
% values, v ::= () | left v | right v | (v,v) 
%
% combinator types, t ::= b <-> b
% primitive combinators, iso ::= zeroe | zeroi | swap+ | assocl+ | assocr+ 
%                   &|& unite | uniti | swap* | assocl* | assocr* 
%                   &|& dist0 | factor0 | dist | factor 
% combinators, c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
\end{definition}

By design, for each combinator {{c : b1 <-> b2}}, there is an
\emph{adjoint} {{c^dagger : b2 <-> b1}} acting in the reverse direction.

\begin{definition}[Adjoints {{c^dagger}}]
 The adjoint of a combinator {{c}} is defined by {{c^dagger}} as
 follows. For primitive combinators {{c}}, {{c^dagger}} is given by
 the inverse in Def.~\ref{def:primitives-langRev}. For the remaining
 combinators, we have:
%subcode{opsem} include main
% id^dagger &=& id
% (sym c)^dagger &=& c
% (c1(;)c2)^dagger &=& c2^dagger (;) c1^dagger
% (c1 (+) c2)^dagger &=& c1^dagger (+) c2^dagger
% (c1 (*) c2)^dagger &=& c1^dagger (*) c2^dagger
which essentially amounts to reversing the order of sequential composition.
\end{definition}

The semantics of {{langRev}} is given using two mutually recursive
interpreters: one going forward and one going backwards. The use of
{{sym}} switches control from one evaluator to the other.

\begin{definition}[Operational Semantics of {{langRev}}]
\label{def:operational-langRev}
Given a program {{c : b1 <-> b2}} in {{langRev}}, we can run it in the
forward direction by supplying it with a value {{ v1 : b1 }} and in
the backwards direction by supplying it with a value {{v2 : b2}}. The
forward evaluator {{evalF}} and the backwards evaluator {{evalB}} are
given below. Since there are no values that have the type {{0}}, the
clauses for the combinators {{zeroe}}, {{zeroi}}, {{dist0}} and
{{factor0}} omit the impossible cases.

%subcode{opsem} include main
%! columnStyle = rcl
% evalF zeroe (right v) &=& v
% evalF zeroi v &=& right v
% evalF swap+ (left v) &=& right v
% evalF swap+ (right v) &=& left v 
% evalF assocl+ (left v) &=& left (left v)
% evalF assocl+ (right (left v)) &=& left (right v)
% evalF assocl+ (right (right v)) &=& right v
% evalF assocr+ (left (left v)) &=& left v
% evalF assocr+ (left (right v)) &=& right (left v)
% evalF assocr+ (right v) &=& right (right v)
% evalF unite ((), v) &=& v 
% evalF uniti v &=& ((), v) 
% evalF swap* (v1, v2) &=& (v2, v1) 
% evalF assocl* (v1, (v2, v3)) &=& ((v1, v2), v3) 
% evalF assocr* ((v1, v2), v3) &=& (v1, (v2, v3)) 
% evalF dist (left v1, v3) &=& left (v1, v3)
% evalF dist (right v2, v3) &=& right (v2, v3)
% evalF factor (left (v1, v3)) &=& (left v1, v3) 
% evalF factor (right (v2, v3)) &=& (right v2, v3)   
% evalF id v &=& v
% evalF (sym c) v &=& evalB c v
% evalF (c1(;)c2) v &=& evalF c2 (evalF c1 v)
% evalF (c1 (+) c2) (left v1) &=& left (evalF c1 v1)
% evalF (c1 (+) c2) (right v2) &=& right (evalF c2 v2)
% evalF (c1 (*) c2) (v1,v2) &=& (evalF c1 v1, evalF c2 v2)

%subcode{opsem} include main
%! columnStyle = rcl
% evalB zeroe v &=& right v 
% evalB zeroi (right v) &=& v
% evalB swap+ (left v) &=& right v
% evalB swap+ (right v) &=& left v
% evalB assocl+ (left (left v)) &=& left v
% evalB assocl+ (left (right v)) &=& right (left v)
% evalB assocl+ (right v) &=& right (right v)
% evalB assocr+ (left v) &=& left (left v)
% evalB assocr+ (right (left v)) &=& left (right v) 
% evalB assocr+ (right (right v)) &=& right v
% evalB unite v &=& ((), v) 
% evalB uniti ((), v) &=& v
% evalB swap* (v1, v2) &=& (v2, v1) 
% evalB assocl* ((v1, v2), v3) &=& (v1, (v2, v3)) 
% evalB assocr* (v1, (v2, v3)) &=& ((v1, v2), v3) 
% evalB dist (left (v1, v3)) &=& (left v1, v3)
% evalB dist (right (v2, v3)) &=& (right v2, v3) 
% evalB factor (left v1, v3) &=& left (v1, v3) 
% evalB factor (right v2, v3) &=& right (v2, v3) 
% evalB id v &=& v
% evalB (sym c) v &=& evalF c v
% evalB (c1(;)c2) v &=& evalB c1 (evalB c2 v)
% evalB (c1 (+) c2) (left v1) &=& left (evalB c1 v1)
% evalB (c1 (+) c2) (right v2) &=& right (evalB c2 v2)
% evalB (c1 (*) c2) (v1,v2) &=& (evalB c1 v1, evalB c2 v2)
\end{definition}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Trace Operators: Intuition}

The language {{langRev}} is not yet universal: one of the missing
ingredients is a form of recursion or iteration. Mathematically
speaking, recursion and iteration can be expressed using categorical
trace
operators~\cite{joyal1996traced,Hasegawa:1997:RCS:645893.671607}. We
will work out the precise categorical connections later: for now we
pursue our operational approach.

\paragraph*{Additive Case.} 
The intuitive idea of trace operators is as follows. We consider first
the additive case. Given a computation {{c:b1+b2<->b1+b3}}, we can
\emph{cancel} the common type {{b1}} from both the input and the
output by feeding back to itself, thereby getting a new combinator
{{trace+~c:b2<->b3}}. Diagrammatically, we have:

\inkscape{trace_plus.pdf}

Operationally, the execution of such a combinator proceeds as
follows. We are given a value {{v2 : b2}} which is injected into the
type {{b1 + b2}} by tagging it as {{right v2}}. This tagged value is
passed to the combinator {{f : b1 + b2 <-> b1 + b3}}. The resulting
value will either be {{left v1}} for some {{v1 : b1}} or {{right v3}}
for some {{v3 : b3}}. In the former case, the value {{left v1}} is fed
back to {{f}}. In the latter case, the entire {{trace+}}-combinator
terminates with value {{v3}}. An example should make this more
concrete.

\begin{example}
Consider the combinator {{c:bool+1<->bool+1}} below:

\inkscape{trace-example0.pdf}

\noindent
When we trace {{c}} we have the circuit {{trace+ c: 1 <-> 1}} shown below.

\inkscape{trace-example1.pdf}

\noindent
The following diagrams show snapshots the evaluation of the
circuit applied to {{()}}:
\begin{multicols}{2}
  \inkscape{trace-example2.pdf}
  \inkscape{trace-example3.pdf}
\end{multicols}

\begin{multicols}{2}
  \inkscape{trace-example4.pdf}
  \inkscape{trace-example5.pdf}
\end{multicols}

\begin{multicols}{2}
  \inkscape{trace-example6.pdf}
  \inkscape{trace-example7.pdf}
\end{multicols}

\noindent
The value starts at the input to the circuit, flows out on the
{{left}} branch of {{c}} and is traced back into {{c}}.  The feedback
loop is taken exactly two times in the execution of this circuit.
\end{example}

\paragraph*{Multiplicative Case.} 
The situation with the multiplicative trace is a bit more subtle. In
this case, we are given a computation {{c:b1 * b2 <-> b1 * b3}} and we
want to cancel the common type {{b1}} as before, to produce a new
combinator {{trace*~c:b2<->b3}}. For the evaluation of such a
combinator, we are only given a value {{v2 : b2}} but to evaluate
{{c}}, we must provide a pair consisting of some value {{v1 : b1}}
together with the given {{v2 : b2}}. The needed value {{v1 : b1}}
cannot be arbitrary, however. We do have a constraint that the value
{{v1 : b1}} must be such that it is also produced by as the first
component of the result, i.e., that {{evalF c (v1,v2) = (v1,v3)}}. In
general, there may be several such values fixed-point values or
none. A few examples should help understand the subtleties.

\begin{example}
One might conclude from the examples above that the multiplicative
trace is non-sensical. There are however some excellent and legitimate
uses illustrated in the remainder of this chapter. A simple
illustrative example takes a combinator {{c : b <-> b}} and detects
whether it has a fixed point, i.e., whether there exists a value 
{{v :  b}} such that {{evalF c v = v}}:

{{hasFixedPoint c = trace* (swap* (;) unite (;) c (;) uniti (;) swap*)}}

\noindent The idea is to introduce a unit value before and after the evaluation of {{c}}
effectively changing the type of {{c}} to {{b * 1 <-> b * 1}}
and then use {{trace*}} to
cancel the value of type~{{b}}. The result of evaluating
{{(hasFixedPoint c) ()}} is {{()}} if {{c}} has a fixed point and returns 
no value otherwise. 
\end{example}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Multiplicative Trace Operators}
\label{sec:trace}

For the multiplicative trace operator, we similarly add a new family
of combinators {{trace*_{b} c}} indexed by types {{b}} to our
syntax. The type rule, can be viewed as a cancellation law that
eliminates the type~{{b}} from both sides of a product.

\begin{definition}[Type Judgment for {{trace*_{b}}}]
%subcode{proof} include main
%@ c : b * b2 <-> b * b3
%@@ trace*_{b} c : b2 <-> b3  
\end{definition}

The adjoint of {{trace*_{b} c}} is also the trace of the adjoint of
{{c}}, i.e., {{(trace*_{b} c)^dagger = trace*_{b} c^dagger}}.

As hinted in the first section, the semantics of {{trace*}} introduces
a form of non-determinism in the sense that there could zero, one, or
several values that satisfy the constraint imposed by {{trace*}}. To
model such non-determinism, we specify the semantics using
\emph{relations}. In other words, the results of {{evalF c v}} 
or {{evalB c v}} are no longer guaranteed to be a single value: they 
could be \emph{sets} of values.

@@

\begin{definition}[Operational Semantics for {{trace*_{b}}}]
The formal semantics is expressed by lifting the forward and backward
evaluators to relations and then adding the new cases for {{trace*}}.

@@

%subcode{opsem} include main
%! columnStyle = rcl
% evalF zeroe (right v) &=& {[ v ]}
% evalF zeroi v &=& {[ right v ]}
% evalF swap+ (left v) &=& {[ right v ]}
% evalF swap+ (right v) &=& {[ left v ]}
% evalF assocl+ (left v) &=& {[ left (left v) ]}
% evalF assocl+ (right (left v)) &=& {[ left (right v) ]}
% evalF assocl+ (right (right v)) &=& {[ right v ]}
% evalF assocr+ (left (left v)) &=& {[ left v ]}
% evalF assocr+ (left (right v)) &=& {[ right (left v) ]}
% evalF assocr+ (right v) &=& {[ right (right v) ]}
% evalF unite ((), v) &=& {[ v ]}
% evalF uniti v &=& {[ ((), v) ]}
% evalF swap* (v1, v2) &=& {[ (v2, v1) ]}
% evalF assocl* (v1, (v2, v3)) &=& {[ ((v1, v2), v3) ]}
% evalF assocr* ((v1, v2), v3) &=& {[ (v1, (v2, v3)) ]}
% evalF dist (left v1, v3) &=& {[ left (v1, v3) ]}
% evalF dist (right v2, v3) &=& {[ right (v2, v3) ]}
% evalF factor (left (v1, v3)) &=& {[ (left v1, v3) ]}
% evalF factor (right (v2, v3)) &=& {[ (right v2, v3) ]}
% evalF id v &=& {[ v ]}
% evalF (sym c) v &=& evalB c v
% evalF (c1(;)c2) v &=& Union ~{[ evalF c2 v' ~|~ v' in (evalF c1 v) ]}
% evalF (c1 (+) c2) (left v1) &=& {[ left v' ~|~ v' in (evalF c1 v1) ]}
% evalF (c1 (+) c2) (right v2) &=& {[ left v' ~|~ v' in (evalF c2 v2) ]}
% evalF (c1 (*) c2) (v1,v2) &=& {[ (v3,v4) ~|~ v3 in (evalF c1 v1), v4 in (evalF c2 v2) ]}
% evalF (trace*_{b} c) v2 &=& Union_{v in [[b]]} ~{[ v3 ~|~ (v,v3) in (evalF c (v,v2)) ]}

@@

%subcode{opsem} include main
%! columnStyle = rcl
% evalB zeroe v &=& {[ right v ]}
% evalB zeroi (right v) &=& {[ v ]}
% evalB swap+ (left v) &=& {[ right v ]}
% evalB swap+ (right v) &=& {[ left v ]}
% evalB assocl+ (left (left v)) &=& {[ left v ]}
% evalB assocl+ (left (right v)) &=& {[ right (left v) ]}
% evalB assocl+ (right v) &=& {[ right (right v) ]}
% evalB assocr+ (left v) &=& {[ left (left v) ]}
% evalB assocr+ (right (left v)) &=& {[ left (right v) ]}
% evalB assocr+ (right (right v)) &=& {[ right v ]}
% evalB unite v &=& {[ ((), v) ]}
% evalB uniti ((), v) &=& {[ v ]}
% evalB swap* (v1, v2) &=& {[ (v2, v1) ]}
% evalB assocl* ((v1, v2), v3) &=& {[ (v1, (v2, v3)) ]}
% evalB assocr* (v1, (v2, v3)) &=& {[ ((v1, v2), v3) ]}
% evalB dist (left (v1, v3)) &=& {[ (left v1, v3) ]}
% evalB dist (right (v2, v3)) &=& {[ (right v2, v3) ]}
% evalB factor (left v1, v3) &=& {[ left (v1, v3) ]}
% evalB factor (right v2, v3) &=& {[ right (v2, v3) ]}
% evalB id v &=& {[ v ]}
% evalB (sym c) v &=& evalF c v
% evalB (c1(;)c2) v &=& Union ~{[ evalB c1 v' ~|~ v' in (evalB c2 v) ]}
% evalB (c1 (+) c2) (left v1) &=& {[ left v' ~|~ v' in (evalB c1 v1) ]}
% evalB (c1 (+) c2) (right v2) &=& {[ right v' ~|~ v' in (evalB c2 v2) ]}
% evalB (c1 (*) c2) (v1,v2) &=& {[ (v3,v4) ~|~ v3 in (evalB c1 v1), v4 in (evalB c2 v2) ]}
% evalB (trace*_{b} c) v2 &=& Union_{v in [[b]]} ~{[ v3 ~|~ (v,v3) in (evalB c (v,v2)) ]}
\end{definition}

In the cases for {{trace*_b}}, we check every value {{v : b}} in the
denotation of the type {{b}}. Specifically, given {{v2}}, we produce the pair
{{(v,v2)}} and check if the result of evaluating {{c}} on this pair produces
the same {{v}} as the first component of its result. In that case, we include
the second component in the final answer, and otherwise, we reject that
second component.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{(Finite) Relational Programming}
\label{ch3:sec:lp}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Solving Constraints}
\label{ch3:sec:constraints}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The Idea}

Our development so far has been fairly traditional in the sense that
we have applied well-known type constructions and semantic techniques
to a language of isomorphisms. In this chapter, we begin to explore
more novel ideas.

One of the benefits of having a reversible language in which
information is preserved is that essentially every construction can be
considered in reverse. We begin in this chapter by considering the
idea of \emph{fractional types} which are dual to product types. We
call the resulting language {{langRevTF}}.

We have already seen in Sec.~\ref{sec:trace} that the multiplicative
trace can be viewed as performing some kind of ``division'' on
types. Recall that given a combinator {{c : b * b1 <-> b * b2}}, we
have {{trace* c : b1 <-> b2}} which \emph{cancels} the type {{b}}.
Our plan is to make this ``division operation'' a first-class one. In
other words, we introduce a new type construction --- a fractional
type {{1/b}} --- that allows us to cancel any type {{b}} in a
multiplicative context, i.e., that satisfies the isomorphism:

%subcode{bnf} include main
%  eps*_b :& 1/b * b ~~<->~~ 1 &: eta*_b

Given this fractional type and its associated isomorphism, we can
perform the cancellation implicit in {{trace*}} as follows. 
Let {{ c :  b * b1 <-> b * b2 }} be given and consider the 
following sequence of type isomorphisms:

%subcode{bnf} include main
%  b1 &<->& 1 * b1 & (uniti)
%     &<->& (1/b * b) * b1 & (eta*_b)
%     &<->& 1/b * (b * b1) & (assocr*)
%     &<->& 1/b * (b * b2) & (c)
%     &<->& ((1/b * b) * b2) & (assocl*)
%     &<->& (1 * b2) & (eps*_b)
%     &<->& b2 & (unite*)

\noindent This sequence of isomorphisms, at least at the type level,
seems to produce the desired {{trace*}}. We will confirm this in the
next section after we introduce the formal syntax and semantics. For
now, we provide an intuitive understanding of the type {{1/b}}.

If {{v : b}}, we define the value {{1/v : 1/b}} and these are the only
values of type~{{1/b}}. Since there are no values of type {{0}}, this
means that {{1/0}} is also an empty type. Generally, we think of a
value like {{1/false}} as a \emph{first-class constraint} that can
only be satisfied if it is matched with an actual value {{false}}. So
intuitively, {{trace*}} allowed us to implicitly introduce constraints
in our model which we now make explicit and first-class entities. This
makes the language much more expressive: as we show in this chapter,
it essentially gains the expressive power to model higher-order
relations.

We extend {{langRevT}} with fractional types, fractional values, and
two combinators {{eta*_b}} and {{eps*_b}} motivated above.  We provide the
full syntactic and semantic definitions for the language below.

The denotation of types is extended with a clause for fractional types.

\begin{definition}[Denotation of Types {{ [[b]] }}]
\label{chx:def:denot}
Each type denotes a set of values as follows:
\[\begin{array}{rcl}
\llbracket 0 \rrbracket &=& \emptyset \\
\llbracket 1 \rrbracket &=& \{ () \} \\
\llbracket b_1 + b_2 \rrbracket &=& \{ \mathit{left}~v ~|~ v \leftarrow \llbracket b_1 \rrbracket \}
           \cup \{ \mathit{right}~v ~|~ v \leftarrow \llbracket b_2 \rrbracket \} \\
\llbracket b_1 \times b_2 \rrbracket &=& \{ (v_1,v_2) ~|~ v_1 \leftarrow \llbracket b_1 \rrbracket, 
           v_2 \leftarrow \llbracket b_2 \rrbracket \} \\
\llbracket 1/v \rrbracket &=& \{ 1/v ~|~ v \leftarrow \llbracket b \rrbracket \} \\
\llbracket \mathit{bool} \rrbracket &=& \{ \mathit{true}, \mathit{false} \} \\
\llbracket \mathit{nat} \rrbracket &=& \{ 0, 1, 2, \ldots \} 
\end{array}\]
\end{definition}

As usual, the semantics of {{langRevTF}} is given using two mutually
recursive interpreters: one going forward and one going backwards. As
motivated and explained in Sec.~\ref{sec:trace}, each combinator denotes
a \emph{relation} and hence the result of applying a combinator to a
value generally returns a \emph{set} of values. We only show the new
cases for {{eta*_b}} and {{eps*_b}}.

\paragraph*{Forward Evaluator (clauses for {{eta*_b}} and {{eps*_b}}): }

%subcode{opsem} include main
%! columnStyle = rcl
% evalF ~eta*_b () &=& Union_{v in [[b]]} ~{[ (1/v,v) ]}
% evalF ~eps*_b (1/v,v) &=& {[ () ]} 
% evalF ~eps*_b (1/v1,v2) &=& emptyset, where v1 /= v2

\paragraph*{Backwards Evaluator (clauses for {{eta*_b}} and {{eps*_b}}): }

%subcode{opsem} include main
%! columnStyle = rcl
% evalB ~eta*_b (1/v,v) &=& {[ () ]} 
% evalB ~eta*_b (1/v1,v2) &=& emptyset, where v1 /= v2
% evalB ~eps*_b () &=& Union_{v in [[b]]} ~{[ (1/v,v) ]}
\end{definition}

@@

%%%%%%%%%%%%
\subsection{Trace}

With fractionals, the trace operator becomes expressible:

>>>>

\newcommand{\todo}[1]{\textbf{TODO:} #1}

%subcode-inline{bnf-inline} name langRev
%! swap+ = \mathit{swap}^+
%! swap* = \mathit{swap}^*
%! dagger =  ^{\dagger}
%! assocl+ = \mathit{assocl}^+
%! assocr+ = \mathit{assocr}^+
%! assocl* = \mathit{assocl}^*
%! assocr* = \mathit{assocr}^*
%! identr* = \mathit{uniti}
%! identl* = \mathit{unite}
%! dist = \mathit{distrib}
%! factor = \mathit{factor}
%! (o) = \fatsemi
%! (;) = \fatsemi
%! (*) = \times
%! (+) = +
%! foldB = fold_B
%! unfoldB = unfold_B
%! foldN = fold_N
%! unfoldN = unfold_N
%! trace+ = \mathit{trace}^{+}
%! trace* = \mathit{trace}^{\times}
%! :-* = \multimap^{\times}
%! :-+ = \multimap^{+}
%! emptyset = \emptyset

%subcode-inline{bnf-inline} regex \{\{(((\}[^\}])|[^\}])*)\}\} name main include langRev
%! [^ = \ulcorner
%! ^] = \urcorner
%! [v = \llcorner
%! v] = \lrcorner
%! eta* = \eta^{\times}
%! eps* = \epsilon^{\times}
%! Union = \bigcup
%! in = \in
%! |-->* = \mapsto^{*}
%! |-->> = \mapsto_{\ggg}
%! |-->let = \mapsto_{let}
%! |--> = \mapsto
%! <--| = \mapsfrom
%! |- = \vdash
%! <=> = \Longleftrightarrow
%! <-> = \leftrightarrow
%! ~> = \leadsto
%! ::= = ::=
%! /= = \neq
%! vi = v_i
%! di = d_i
%! si = s_i
%! sj = s_j
%! F = \texttt{F}
%! T = \texttt{T}
%! forall = \forall
%! exists = \exists
%! empty = \emptyset
%! eta = \eta
%! where = \textbf{where}
%! epsilon = \varepsilon
%! least = \phi
%! loop+ = loop_{+}
%! loop* = loop_{\times}
%! CatC = {\mathcal C}
%! CatA = {\mathcal A}
%! gamma = \gamma
%! {[ = \{
%! ]} = \}
%! elem = \in
%! dagger = ^\dagger
%! alpha = \alpha
%! beta = \beta
%! rho = \rho
%! @@ = \mu
%! @ = \,@\,
%! langRev = \Pi
%! langRevT = \Pi^{o}
%! langRevEE = \Pi^{\eta\epsilon}_{+}
%! bullet = \bullet
%! * = \times
%! langRevT = \Pi^{o}
%! langRevTF = \Pi^{o/}

@@

\begin{document}
\title{Fractional Types}
\author{Roshan P. James \and Zachary Sparks \and Jacques Carette \and Amr Sabry}
\institute{}
\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
  In previous work, we developed a \emph{first-order},
  information-preserving, and reversible programming language {{langRev}}
  founded on type isomorphisms. Being first-order limits the expressiveness
  of the language: it is not possible, for example, to abstract program
  fragments into a higher-level combinator. More importantly, however, the
  limitation to first-order types means that the class of isomorphisms
  between two types {{b1}} and {{b2}} cannot be itself represented as a type
  and hence that the natural notion of equivalence between two {{langRev}}
  computations is no longer based on isomorphisms. In this paper, we
  introduce a higher-order extension of {{langRev}} based on the novel
  concept of \emph{fractional types}. Intuitively, a value of a fractional
  type represents \emph{negative} information: such fractional values can be
  freely propagated and transformed but must ultimately --- in a complete
  program --- be offset by the corresponding amount of positive information.
\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Background: {{langRev}} }

\todo{brief explanation}

@@

The semantics of {{langRev}} is given using two mutually recursive
interpreters: one going forward and one going backwards. The use of {{sym}}
switches control from one evaluator to the other. We will not present the
semantics because we generalize in the next section.

@@

\todo{mention that we have extended this with trace and recursive types but
  that we are going to explore something orthogonal}

As hinted to in the first section, the semantics of fractional types introduces
a form of non-determinism in the sense that there could zero, one, or several
values that satisfy the constraints imposed by fractional values. To model
such non-determinism, we specify the semantics using \emph{relations}. In
other words, the results of {{evalF c v}} or {{evalB c v}} are no longer
guaranteed to be a single value: they could be \emph{sets} of values.

@@

We extend {{langRevT}} with fractional types, fractional values, and two
combinators {{eta*_b}} and {{eps*_b}}.  We provide the full syntactic and
semantic definitions for the language below.

If {{v : b}}, we define the value {{1/v : 1/b}} and these are the only
values of type~{{1/b}}. Since there are no values of type {{0}}, this
means that {{1/0}} is also an empty type. Generally, we think of a
value like {{1/false}} as a \emph{first-class constraint} that can
only be satisfied if it is matched with an actual value {{false}}.

\begin{definition}[Denotation of Types {{ [[b]] }}]
\label{chx:def:denot}
Each type denotes a set of values as follows:
\[\begin{array}{rcl}
\llbracket 0 \rrbracket &=& \emptyset \\
\llbracket 1 \rrbracket &=& \{ () \} \\
\llbracket b_1 + b_2 \rrbracket &=& \{ \mathit{left}~v ~|~ v \leftarrow \llbracket b_1 \rrbracket \}
           \cup \{ \mathit{right}~v ~|~ v \leftarrow \llbracket b_2 \rrbracket \} \\
\llbracket b_1 \times b_2 \rrbracket &=& \{ (v_1,v_2) ~|~ v_1 \leftarrow \llbracket b_1 \rrbracket, 
           v_2 \leftarrow \llbracket b_2 \rrbracket \} \\
\llbracket 1/b \rrbracket &=& \{ 1/v ~|~ v \leftarrow \llbracket b \rrbracket \} \\
\llbracket \mathit{bool} \rrbracket &=& \{ \mathit{true}, \mathit{false} \} \\
\llbracket \mathit{nat} \rrbracket &=& \{ 0, 1, 2, \ldots \} 
\end{array}\]
\end{definition}

The semantics of {{langRevTF}} is given using two mutually recursive
interpreters: one going forward and one going backwards. Unlike the case for
{{langRev}}, each combinator denotes a \emph{relation} and hence the result
of applying a combinator to a value generally returns a \emph{set} of
values.

@@

%subcode{opsem} include main
%! columnStyle = rcl
% evalF zeroe (right v) &=& {[ v ]}
% evalF zeroi v &=& {[ right v ]}
% evalF swap+ (left v) &=& {[ right v ]}
% evalF swap+ (right v) &=& {[ left v ]}
% evalF assocl+ (left v) &=& {[ left (left v) ]}
% evalF assocl+ (right (left v)) &=& {[ left (right v) ]}
% evalF assocl+ (right (right v)) &=& {[ right v ]}
% evalF assocr+ (left (left v)) &=& {[ left v ]}
% evalF assocr+ (left (right v)) &=& {[ right (left v) ]}
% evalF assocr+ (right v) &=& {[ right (right v) ]}
% evalF unite ((), v) &=& {[ v ]}
% evalF uniti v &=& {[ ((), v) ]}
% evalF swap* (v1, v2) &=& {[ (v2, v1) ]}
% evalF assocl* (v1, (v2, v3)) &=& {[ ((v1, v2), v3) ]}
% evalF assocr* ((v1, v2), v3) &=& {[ (v1, (v2, v3)) ]}
% evalF dist (left v1, v3) &=& {[ left (v1, v3) ]}
% evalF dist (right v2, v3) &=& {[ right (v2, v3) ]}
% evalF factor (left (v1, v3)) &=& {[ (left v1, v3) ]}
% evalF factor (right (v2, v3)) &=& {[ (right v2, v3) ]}
% evalF id v &=& {[ v ]}
% evalF (sym c) v &=& evalB c v
% evalF (c1(;)c2) v &=& Union ~{[ evalF c2 v' ~|~ v' in (evalF c1 v) ]}
% evalF (c1 (+) c2) (left v1) &=& {[ left v' ~|~ v' in (evalF c1 v1) ]}
% evalF (c1 (+) c2) (right v2) &=& {[ left v' ~|~ v' in (evalF c2 v2) ]}
% evalF (c1 (*) c2) (v1,v2) &=& {[ (v3,v4) ~|~ v3 in (evalF c1 v1), v4 in (evalF c2 v2) ]}
% evalF (trace*_{b} c) v2 &=& Union_{v in [[b]]} ~{[ v3 ~|~ (v,v3) in (evalF c (v,v2)) ]}
% evalF ~eta*_b () &=& Union_{v in [[b]]} ~{[ (1/v,v) ]}
% evalF ~eps*_b (1/v,v) &=& {[ () ]} 
% evalF ~eps*_b (1/v1,v2) &=& emptyset, where v1 /= v2

@@

%subcode{opsem} include main
%! columnStyle = rcl
% evalB zeroe v &=& {[ right v ]}
% evalB zeroi (right v) &=& {[ v ]}
% evalB swap+ (left v) &=& {[ right v ]}
% evalB swap+ (right v) &=& {[ left v ]}
% evalB assocl+ (left (left v)) &=& {[ left v ]}
% evalB assocl+ (left (right v)) &=& {[ right (left v) ]}
% evalB assocl+ (right v) &=& {[ right (right v) ]}
% evalB assocr+ (left v) &=& {[ left (left v) ]}
% evalB assocr+ (right (left v)) &=& {[ left (right v) ]}
% evalB assocr+ (right (right v)) &=& {[ right v ]}
% evalB unite v &=& {[ ((), v) ]}
% evalB uniti ((), v) &=& {[ v ]}
% evalB swap* (v1, v2) &=& {[ (v2, v1) ]}
% evalB assocl* ((v1, v2), v3) &=& {[ (v1, (v2, v3)) ]}
% evalB assocr* (v1, (v2, v3)) &=& {[ ((v1, v2), v3) ]}
% evalB dist (left (v1, v3)) &=& {[ (left v1, v3) ]}
% evalB dist (right (v2, v3)) &=& {[ (right v2, v3) ]}
% evalB factor (left v1, v3) &=& {[ left (v1, v3) ]}
% evalB factor (right v2, v3) &=& {[ right (v2, v3) ]}
% evalB id v &=& {[ v ]}
% evalB (sym c) v &=& evalF c v
% evalB (c1(;)c2) v &=& Union ~{[ evalB c1 v' ~|~ v' in (evalB c2 v) ]}
% evalB (c1 (+) c2) (left v1) &=& {[ left v' ~|~ v' in (evalB c1 v1) ]}
% evalB (c1 (+) c2) (right v2) &=& {[ right v' ~|~ v' in (evalB c2 v2) ]}
% evalB (c1 (*) c2) (v1,v2) &=& {[ (v3,v4) ~|~ v3 in (evalB c1 v1), v4 in (evalB c2 v2) ]}
% evalB (trace*_{b} c) v2 &=& Union_{v in [[b]]} ~{[ v3 ~|~ (v,v3) in (evalB c (v,v2)) ]}
% evalB ~eta*_b (1/v,v) &=& {[ () ]} 
% evalB ~eta*_b (1/v1,v2) &=& emptyset, where v1 /= v2
% evalB ~eps*_b () &=& Union_{v in [[b]]} ~{[ (1/v,v) ]}
\end{definition}

@@

@@
